Design of 1Mbit RRAM memory for NV applications

### 2014-05-23 Thesis/Latex/hfdst-sensamp.tex
sensitivity lower Vdd

Men kan argumenteren dat er een korte overlap zou kunnen toegelaten zijn, waarna er voldoende spanningsverschil tussen de 2 ingangs-uitgangsknopen zou opgebouwd zijn opdat de ladingsinjectie geen effect meer kan hebben op het eindresultaat. Een tegenargument is dat de timing hiervoor te precies moet zijn.
\begin{figure}
  \centering
  \includegraphics[scale=0.4]{../fig/hfdstk-sensamp-chargeinjectionmismatch.png}
  \caption[foutief latchen door $\beta$-mismatch]{Door $\beta$-mismatch is ladingsinjectie van de passgates niet meer gematched en gaat de SA foutief latchen}
  \label{fig:chargeinjectionmismatch}
\end{figure}

>>>>

Men kan argumenteren dat er een korte overlap zou kunnen toegelaten zijn, waarna er voldoende spanningsverschil tussen de 2 ingangs-uitgangsknopen zou opgebouwd zijn opdat de ladingsinjectie geen effect meer kan hebben op het eindresultaat. Een tegenargument is dat de timing hiervoor te precies moet zijn.
\begin{figure}
  \centering
  \includegraphics[scale=0.4]{../fig/hfdstk-sensamp-chargeinjectionmismatch.png}
  \caption[Foutief latchen door $\beta$-mismatch]{Door $\beta$-mismatch is ladingsinjectie van de passgates niet meer gematched en gaat de SA foutief latchen}
  \label{fig:chargeinjectionmismatch}
\end{figure}

### 2014-05-23 Thesis/Latex/masterproef.tex
sensitivity lower Vdd

\begin{preface}
  Na bijna een jaar aan deze thesis gewerkt te hebben zouden we graag een aantal mensen in het bijzonder willen bedanken voor hun hulp en steun. Ten eerste Prof. Dehaene voor het onderwerp aan te bieden en toe te wijzen aan ons. Ten tweede onze begeleiders Stefan en Burak.\\
  Stefan, hartelijk dank voor de tijd die je wou vrij maken voor ons, het was heel aangenaam om een begeleider te hebben die zo veel kennis en ervaring heeft van het vakgebied.\\
  Burak, thanks for always being available for us and for providing us the resources to get our thesis underway.\\
  Verder zouden we graag Bert DeKnuydt willen bedanken voor zijn hulp met het Condor systeem. Tenslotte zouden we ook elkaar willen bedanken, we waren een goed complemenair team. 
\end{preface}

@@

\begin{abstract}
 Nu het schalen van flash-geheugens op zijn limieten begint te stoten, is er nood aan een alternatief. Met eigenschappen zoals lage voedingsspanning, kleine geheugencel en snelle leessnelheid is RRAM één van de meest belovende kandidaten.\\\\
 Deze thesis behandelt het ontwerp van het leescircuit van een 1Mbit RRAM geheugen. De grootste uitdagingen in het ontwerp van dit resistief geheugen is variabiliteit. Deze variabiliteit is kritisch op twee plaatsen. Ten eerste bij de resistieve deling tussen de geheugencel en een last. Ten tweede bij het latchen van de sense amplifier.\\\\
 Dit werk introduceert verschillende innovaties. Ten eerste een uitgebreide analyse op de keuze van lastimpedantie. Ten tweede het vormen van de referentiespanningsdistributie door middel van parallelle geheugencellen. En tenslotte een analyse over de invloed van een overlappende werking van pass-gates en sense amplifier onder variabiliteit. Verder wordt ook het ontwerp van alle andere belangrijke bouwblokken zoals decoders en buffers besproken.\\\\
 Deze kennis wordt dan gebruikt in het ontwerp van een 1Mbit geheugen in 45nm PTM technologie. Het geheugen maakt gebruik van 512 sense ampliers die telkens gekoppeld zijn aan 2 geheugenmatrices met 32WL en 32BL. Op een voedingspanning van 1V heeft het geheugen een random-access-leessnelheid van 435MHz. Het energieverbruik per leesoperatie is bij deze voedingspanning 0.51pJ. Binnen de afbakening van dit werk presteert de ontworpen schakeling beter dan flash-geheugens gevonden in de literatuur.
\end{abstract}

>>>>

\begin{preface}
  Na bijna een jaar aan deze thesis gewerkt te hebben zouden we graag een aantal mensen in het bijzonder willen bedanken voor hun hulp en steun. Ten eerste prof. Dehaene voor het onderwerp aan te bieden en toe te wijzen aan ons. Ten tweede onze begeleiders Stefan en Burak.\\
  Stefan, hartelijk dank voor de tijd die je wou vrij maken voor ons, het was heel aangenaam om een begeleider te hebben die zo veel kennis en ervaring heeft van het vakgebied.\\
  Burak, thanks for always being available for us and for providing us the resources to get our thesis underway.\\
  Verder zouden we graag Bert DeKnuydt willen bedanken voor zijn hulp met het Condor systeem. Tenslotte zouden we ook elkaar willen bedanken, we waren een goed complemenair team. 
\end{preface}

@@

\begin{abstract}
 Nu het schalen van flash-geheugens op zijn limieten begint te stoten, is er nood aan een alternatief. Met eigenschappen zoals lage voedingspanning, kleine geheugencel en hoge leessnelheid is RRAM één van de meest belovende kandidaten.\\\\
 Deze thesis behandelt het ontwerp van het leescircuit van een 1Mbit RRAM geheugen. De grootste uitdaging in het ontwerp van dit resistief geheugen is variabiliteit. Deze variabiliteit is kritisch op twee plaatsen: ten eerste bij de resistieve deling tussen de geheugencel en een lastimpedantie. Ten tweede bij het latchen van de sense amplifier.\\\\
 Dit werk introduceert verschillende innovaties. Zo werd er een uitgebreide analyse uitgevoerd voor de keuze van lastimpedantie. Het vormen van de referentiespanningsdistributie door middel van parallelle geheugencellen is ook een toegepaste techniek. Tenslotte wordt een analyse over de invloed van een overlappende werking van pass-gates en sense amplifier onder variabiliteit aangebracht. Verder wordt ook het ontwerp van alle andere belangrijke bouwblokken zoals decoders en buffers besproken.\\\\
 Deze kennis wordt dan gebruikt in het ontwerp van een 1Mbit geheugen in 45nm PTM technologie. Het geheugen maakt gebruik van 512 sense amplifiers die telkens gekoppeld zijn aan 2 geheugenmatrices met 32WL en 32BL. Op een voedingspanning van 1V heeft het geheugen een random-access-leessnelheid van 435MHz. Het energieverbruik per leesoperatie is bij deze voedingspanning 0.51pJ. Binnen de afbakening van dit werk presteert de ontworpen schakeling beter dan flash-geheugens gevonden in de literatuur.
\end{abstract}

### 2014-05-23 Thesis/Paper/Paper_template.tex
sensitivity lower Vdd

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Write abstract
A 1Mbit RRAM memory is presented. The focus lies on read operation, write operation has not been included in the design. Reference signal distribution can be modified by connecting reference cells in parallel. Sense amplifier performance has been improved by allowing overlap between passgate-enable and latch-enable signals, this overlap gives rise to a nonlinear phenomenon, christened the RC-latch-effect. All Spectre simulations have been performed with 45nm PTM transistor models.

@@

\section{Load analysis}
Due to the aforementioned variations in the circuit, the data and reference signals should be designed as such that they are sufficiently far apart. After all the sense amplifier will have an offset voltage because of these same variations and the difference of its inputs must be larger than this offset in order for the SA to latch correctly. Besides designing the SA to have a small offset spread to realize this, the difference of its inputs can be widened by choosing a good load impedance. The load impedance not only influences the value of the data and reference voltages, it also determines the settling time of the charging of the BL and the voltage drop over the memristor. This drop can not be too high, destructive reads might occur because of this. As it turns out, fast settling is not compatible with low memristor voltage drop and large voltage difference. Because the latter are imperative for a functional memory and the former is not, settling time had no bearing on the final choice of load impedance.

@@

\paragraph{}
The overlap between the enabling of the passgates and the SA thus needn't be this great: a load-less SA latching delay suffices. For this situation $V_{T}$ and $\beta$ mismatch of the ;passgate transistors results in mismatch of the resistors. The interaction between these resistors and the SA is strongly non-linear. Depending on the precharged values of the load capacitances, this R mismatch can result in incorrect latching. There is thus still a contribution of the passgates to the offset voltage spread. The interaction between R and the SA reduces the contributions of the differential pair mismatch however. For a minimal SA, the offset voltage spread is smaller with overlap than without. By sizing all the transistors, this spread can be arbitrarily reduced. The resistance of the passgates mustn't be too small however, or slow settling might occur. Then again, a large SA would probably charge the load capacitance effortlessly anyway.
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results.}
%\label{fig_sim}
%\end{figure}

@@

\section{Conclusion}
\cite{h2g2}

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Write abstract
A 1Mbit RRAM memory is presented. The focus lies on read operation, write operation has not been included in the design. Reference signal distribution can be modified by connecting reference cells in parallel. A bitline load has been chosen for maximum bitline voltage difference. Sense amplifier performance has been improved by allowing overlap between passgate-enable and latch-enable signals, this overlap gives rise to a nonlinear phenomenon, christened the RC-latch-effect. All Spectre simulations have been performed with 45nm PTM transistor models.

@@

\section{Load analysis}
Due to the aforementioned variations in the circuit, the data and reference signals should be designed as such that they are sufficiently far apart. After all the sense amplifier will have an offset voltage because of these same variations and the difference of its inputs must be larger than this offset in order for the SA to latch correctly. Besides designing the SA to have a small offset spread to realize this, the difference of its inputs can be widened by choosing a good load impedance. The load impedance not only influences the value of the data and reference voltages, it also determines the settling time of the charging of the BL and the voltage drop over the memristor. This drop can not be too high, destructive reads might occur because of this. As it turns out, fast settling is not compatible with low memristor voltage drop and large voltage difference. Because the latter are imperative for a functional memory and the former is not, settling time had no bearing on the final choice of load impedance.
Any type of load consisting of minimal length transistors could not produce an equivalent resistance large enough to separate the HRS \& LRS voltage distributions sufficiently. Figure \ref{fig:switchloadresults} shows nominal HRS/LRS voltage differences and memristor voltage drop for several pMOS transistor loads with different sizes.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/hfdst-last-length.png}
  \caption{HRS/LRS voltage difference and memristor voltage drop for pMOS transistor loads}
  \label{fig:switchloadresults}
\end{figure}

In the end a transistor with a width of 300nm and a length of 198nm was chosen. In figure \ref{fig:distributions} the distributions of the data and reference voltages are displayed for this load.

@@

\paragraph{}
The overlap between the enabling of the passgates and the SA thus needn't be this great: a load-less SA latching delay suffices. For this situation $V_{T}$ and $\beta$ mismatch of the ;passgate transistors results in mismatch of the resistors. The interaction between these resistors and the SA is strongly non-linear. Depending on the precharged values of the load capacitances, this R mismatch can result in incorrect latching. There is thus still a contribution of the passgates to the offset voltage spread. The interaction between R and the SA reduces the contributions of the differential pair mismatch however. For a minimal SA, the offset voltage spread is smaller with overlap than without. By sizing all the transistors, this spread can be arbitrarily reduced. The resistance of the passgates mustn't be too small however, or slow settling might occur. Then again, a large SA would probably charge the load capacitance effortlessly anyway.

\section{Read Throughput Results}
A read access time simulation test was performed on the finished circuit (Figure \ref{fig:speedvdd}). In this test the supply voltage was decreased and the maximum read throughput was measured. The test was performed at a (SPICE) temperature of $30^{\circ}\mathrm{C}$. At each point in the shmoo plot, 100 MonteCarlo simulations were performed. The circuit has a read access time of 2.3ns at a supply voltage of 1V. during the read cycle, its energy consumption is 0.51pJ. Most of the energy (65\%) is consumed by the bitlines. 25\% of the energy consumption is due to the logic, the buffers and SA take the remaining 10\%. As the supply voltage goes down, the read throughput also goes down. This is caused by a combination of phenomena. First of all the logic operates slower. This causes a timing issue in which the bitline load transistor is activated before the the cell is. This timing issue results in a rapid increase in bitline voltage which causes a longer settling time of the bitline when the cell is eventually selected. A second reason for the decrease in read throughput at a lower supply voltage, is the slower latching of the SA.

\begin{figure}[h!t]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/hfdst-final-vddspeed.png}
  \caption{Results read access time test}
  \label{fig:speedvdd}
\end{figure}

On figure \ref{fig:speedvdd} it can clearly be seen that the circuit can only operate down to a supply voltage of $0.8V$. An explanation for this is given in figure \ref{fig:vblvdd}. As vdd is decreased, the bitline voltage distributions of the reference signal and HRS \& LRS data signals also decrease. Eventually the bitline voltage distributions overlap and there is a high probability of circuit failure.

\begin{figure}[h!t]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/hfdst-final-vddbl.png}
  \caption{BL voltage distribution for reference signal and HRS \& LRS signals in function of vdd}
  \label{fig:vblvdd}
\end{figure}
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results.}
%\label{fig_sim}
%\end{figure}

@@

\section{Conclusion}
A 1Mb RRAM memory has been designed and presented. During the design of this memory, certain techniques were applied to improve performance and reliability. Tuning of the reference voltage distribution was achieved by connecting multiple cells in various states in parallel. A load impedance was carefully chosen to maximize HRS \& LRS voltage differences while keeping the memristor voltage drop sufficiently low. Overlap between passgates and sense amplifier operation reduces offset voltage spread. During this overlap, an RC-latch effect can occur, which decouples the bitline capacitance from the latch for a short time.
